---
title: "Lecture 4.2"
---

```{r}
# load packages ----
library(metajam)
library(tidyverse)

# download data from DataOne ----
download_d1_data("https://cn.dataone.org/cn/v2/resolve/https%3A%2F%2Fpasta.lternet.edu%2Fpackage%2Fdata%2Feml%2Fknb-lter-hbr%2F208%2F9%2F3b3cf7ea447cb875d7c7d68ebdfd24c7",
                 path = here::here("week4"))

# read in downloaded files ----
stream_chem_all <- read_d1_files(here::here("week4", "data"))

# get the data file ----
stream_chem_data <- stream_chem_all$data
```
```{r}
# basic scatter ----
stream_chem_data %>% 
  filter(waterYr == 2021) %>% 
  ggplot(aes(x = DOC, y = pH)) +
  geom_point(alpha = 0.5) +
  geom_rug()

# ggExtra ----
p1 <- stream_chem_data %>% 
  filter(waterYr == 2021) %>% 
  ggplot(aes(x = DOC, y = pH)) +
  geom_point(alpha = 0.5)

ggExtra::ggMarginal(p1, type = "density", margins = "x")

# ggExtra multiple groups ----
p2 <- stream_chem_data %>% 
  filter(waterYr == 2021) %>% 
  ggplot(aes(x = DOC, y = pH, color = site)) +
  geom_point(alpha = 0.5) +
  theme(legend.position = "bottom")

ggExtra::ggMarginal(p2, type = "density", groupFill = TRUE, groupColour = TRUE)
```

```{r}
stream_chem_data %>% 
  filter(waterYr == 2021) %>% 
  filter(site == "W8") %>% 
  ggplot(aes(x = DOC, y = pH)) +
  geom_point(alpha = 0.5) +
  geom_smooth(method = "lm", se = FALSE)
```
```{r}
stream_chem_data %>% 
  filter(waterYr == 2021) %>% 
  ggplot(aes(x = DOC, y = pH, color = site, size = Al_ICP)) +
  geom_point(alpha = 0.5) +
  scale_size(range(c(1,10)))
# always scale bubble by area!!
# do not scale by radius because it can be deceiving
```
### Using size to represent a third numeric varaiable cautiously

Bubble charts encode values based on both position (within cartesian coordinate system) *and* size. Some challenges:

- we (humans) generally have an easier time perceiving differences in position rather than size – the relationship between your x- and y-axis variables will be the primary focus
- it can be hard to compare the strengths of different associations – is there an alternative (e.g. 2+ separate scatter plots) way of presenting your data to better highlight this?
- it's difficult to see small differences in size - if the range of values mapped to size is small, your bubles iwll look indistinguishable from one another
- it can be difficult to match a bubble’s size to the scale of difference in the legend – consider adjusting the size range of your scale

## Overplotting

```{r}
ggplot(stream_chem_data, aes(x = SO4, y = pH, color = site)) +
  geom_point(size = 0.5, alpha = 0.5) +
  geom_rug()
```
## heat maps/ density plots

```{r}
# heatmap ----
ggplot(stream_chem_data, aes(x = SO4, y = pH)) +
  geom_bin2d() +
  scale_fill_viridis_c()

# hex map ----
ggplot(stream_chem_data, aes(x = SO4, y = pH)) +
  geom_hex() +
  scale_fill_viridis_c() +
  guides(fill = guide_colorbar(title = "Count",
                               barwidth = 1,
                               barheight = 15))
```

