---
title: "Lecutre 3.3"
---

## setup
```{r}
library(tidycensus)
library(tidyverse)
library(janitor)
library(gghighlight)

source(here::here("week3", "KEYS.R"))
census_api_key(censusKEY)

lyme <- read_csv(here::here("week3", "data", "LD-Case-Counts-by-County-01-20.csv"))
```
## wrangle
```{r}
# lyme data----
lyme_clean <- lyme %>% 
  clean_names() %>% 
  rename(county = ctyname, state = stname, status = ststatus) %>% 
  pivot_longer(cols = 6:25, 
               names_to = "county_year",
               values_to = "reported_cases") %>% 
  mutate(year = str_remove(county_year, pattern = "cases"),
         year = as.factor(year))

# calculate cases by state ----
lyme_by_state <- lyme_clean %>% 
  group_by(year, state) %>% 
  summarize(total_cases = sum(reported_cases))


# grab population data ----
us_state_pop <- get_estimates(geography = "state",
                              product = "population",
                              year = 2019) %>% 
  filter(variable == "POP") %>% 
  select(state = NAME, population = value)

# join data ----
lyme_pop <- left_join(lyme_by_state, us_state_pop) %>% 
  mutate(pop100k = population/100000) %>% 
  mutate(cases_per100k = total_cases/pop100k)
```

## line plot

```{r}
# highlight one group ----
lyme_pop %>% 
  filter(year %in% c(2010:2020)) %>% 
  ggplot(aes(x = year, 
             y = cases_per100k, 
             group = state, 
             color = state)) +
  geom_line() +
  gghighlight::gghighlight(state == "New Jersey")


# highlight based on a conditional ----
lyme_pop %>% 
  filter(year %in% c(2010:2020)) %>% 
  ggplot(aes(x = year, 
             y = cases_per100k, 
             group = state, 
             color = state)) +
  geom_line() +
  gghighlight::gghighlight(max(cases_per100k) > 100)
```
Line graphs encode data by position and not length, therefore, you can choose to include the 0 origin only if it makes sense. 

- Do *not* start the y-axis at 0 if the range of data is small but the distance from the bottom of the range to zero is large.
- Do not start the y-axis at 0 if the relationship to zero is insignificant.
  - ex: human body temperature change over time
  
### Aspect ratio

**Aspect ratio**: height:width ratio of a graph. Manipulating aspect ratio can mislead readers, so do so carefully.

```{r}
lyme_pop %>% 
  filter(year %in% c(2010:2020)) %>% 
  filter(state == "Vermont") %>% 
  ggplot(aes(x = year, y = cases_per100k, group = state)) +
  geom_line() +
  #scale_y_continuous(breaks = seq(0,190, by = 1)) +
  # change aspect ratio of plot
  coord_fixed(ratio = 1/50)
```
## Area plots

```{r}
# area chart one group ----
lyme_pop %>% 
  filter(year %in% c(2010:2020)) %>% 
  filter(state == "Rhode Island") %>% 
  ggplot(aes(x = year, y = cases_per100k, group = state, fill = state)) +
  geom_area() +
  scale_fill_manual(values = c("aquamarine")) +
  scale_x_discrete(expand = c(0, 0)) +
  scale_y_continuous(expand = c(0,0))
  theme(legend.position = "none") +
  labs(title = "Cases of Lyme disease in Rhode Island")
  
# stacked area chart ----
lyme_pop %>% 
  filter(year %in% c(2010:2020)) %>% 
  filter(state %in% c("Maine", "Rhode Island", "New Hampshire", "Vermont")) %>% 
  ggplot(aes(x = year, y = cases_per100k, group = state, fill = state)) +
  geom_area()
## cannot see absolute values, but top line shows sum of all groups
  
# proportional stacked area chart ----
lyme_pop %>% 
  filter(year %in% c(2010:2020)) %>% 
  filter(state %in% c("Maine", "Rhode Island", "New Hampshire", "Vermont")) %>% 
  ggplot(aes(x = year, y = cases_per100k, group = state, fill = state)) +
  geom_area(position = "fill") +
  scale_y_continuous(label = scales::label_percent(scale = 100))
  
# line chart version
lyme_pop |> 
  filter(year %in% c(2010:2020)) |> 
  filter(state %in% c("Maine", "Rhode Island", "New Hampshire", "Vermont")) |> 
  ggplot(aes(x = year, y = cases_per100k, group = state, color = state)) +
  geom_line() +
  facet_wrap(~state)
```
**Plot most important group on bottom so you can read the y-axis**

- to reverse order of states: `geom_area(position = position_fill(reverse = TRUE))`
